Generalized spaces for constructive algebra

### 2020-04-05 paper-herrsching2019/main.tex
Frobenius rules required in geometric logic
Baby Barr (with variables)

% NOTE TO EDITORS AND LAYOUTERS: This package increases the quality of the
% typesetting. I personally recommebd using it, but my contribution does not
% depend on it.
\usepackage[expansion=true,protrusion=true]{microtype}

@@

@@

\paragraph{The geometric theory of rings.} This theory has one sort, $R$; five function symbols:
$0$ and $1$ (nullary), $-$ (unary), $+$ and $\cdot$ (binary); no relation
symbols; and the usual axioms, such as $\top \vdash_{x:R,y:R} x + y = y + x$.

@@

In a similar vein, there is the geometric theory of monoids, of groups, and so
on.

\paragraph{The geometric theory of objects.} This theory has one sort, $X$, and no function
symbols, relation symbols or axioms. A set-based model of this theory is just a
set.

@@

\paragraph{The geometric theory of surjections~$f : \NN \twoheadrightarrow
\RR$.} This theory has no
sorts; no function symbols; a nullary relation symbol~$\varphi_{nx}$ for each
pair~$\langle n,x \rangle \in \NN \times \RR$ (to be read as~``$f$ maps~$n$
to~$x$''); and the following axioms:
\begin{enumerate}
\item For each~$n \in \NN$, the axiom~$\top \vdash \bigvee_{x \in \RR}
\varphi_{nx}$, expressing that~$f$ is total.
\item For each~$n \in \NN$ and each~$x,y \in \RR$, the axiom~$\varphi_{nx}
\wedge \varphi_{ny} \vdash \bigvee\{ \top \,|\, x = y \}$,
expressing that~$f$ is single-valued.\footnote{The disjunction is taken over
the set~$\{ \top \,|\, x = y \}$. This is a certain subsingleton set of
formulas; it is inhabited (by the formula~$\top$) iff~$x = y$. Hence, if~$x =
y$, this axiom reads~$\varphi_{nx}
\wedge \varphi_{ny} \vdash \top$ and could also be omitted; if~$x \neq y$,
this axiom reads~$\varphi_{nx} \wedge \varphi_{ny} \vdash \bot$.}
\item For each~$x \in \RR$, the axiom~$\top \vdash \bigvee_{n \in \NN}
\varphi_{nx}$, expressing that~$f$ is surjective.
\end{enumerate}

@@

\paragraph{The geometric theory of Dedekind cuts.}\label{par:dedekind-cuts}
This theory has no sorts; no
function symbols; nullary relation symbols~$\alpha_x$ and~$\beta_x$ for each
rational number~$x$ (to be read as~``$x$ is contained in the lower (upper)
cut''); and the following axioms, spelling out that the cut is bounded,
rounded, open and located:
\begin{align*}
  \top &\vdash \textstyle\bigvee_{x \in \QQ} \alpha_x \\
  \top &\vdash \textstyle\bigvee_{x \in \QQ} \beta_x \\
  \alpha_y &\vdash \alpha_x & \text{(for each~$x,y \in \QQ$ such that~$x < y$)} \\
  \beta_x &\vdash \beta_y & \text{(for each~$x,y \in \QQ$ such that~$x < y$)} \\
  \alpha_x &\vdash \textstyle\bigvee_{y > x} \alpha_y & \text{(for each~$x \in \QQ$)} \\
  \beta_y &\vdash \textstyle\bigvee_{x < y} \beta_x & \text{(for each~$y \in \QQ$)} \\
  \alpha_x \wedge \beta_y &\vdash \bot & \text{(for each~$x,y \in \QQ$ such that~$x \geq y$)} \\
  \top &\vdash \alpha_x \vee \beta_y & \text{(for each~$x,y \in \QQ$ such that~$x < y$)}
\end{align*}

@@

\paragraph{The geometric theory of prime ideals of a given ring~$A$.} In classical
commutative algebra, the notion of a \emph{prime ideal} of a ring~$A$ is
fundamental. Corresponding to the definition (recalled in
Section~\ref{sect:algebraic-preliminaries}), the geometric theory of prime
ideals has no sorts; no function
symbols; one relation symbol~``$V(x)$'' for each element~$x \in A$; and the
following axioms:
\begin{align*}
  \top &\vdash V(0) \\
  V(x) \wedge V(y) &\vdash V(x+y) & \text{(for each~$x,y \in A$)} \\
  V(x) &\vdash V(xy) & \text{(for each~$x,y \in A$)} \\
  V(1) &\vdash \bot \\
  V(xy) &\vdash V(x) \vee V(y) & \text{(for each~$x,y \in A$)}
\end{align*}

@@

\paragraph{The geometric theory of automorphisms of a field.}\label{par:theory-automorphisms}
Let~$L|k$ be a field
extension. The classifying locale of automorphisms~$L \to L$ which fix~$k$ has
one nullary relation symbol~``$\sigma_{xy}$'' for each pair~$\langle x,y
\rangle \in L$ and the following axioms:
\begin{align*}
  \top &\vdash \bigvee_y \sigma_{xy} & \text{(for each~$x \in L$)} \\
  \top &\vdash \bigvee_x \sigma_{xy} & \text{(for each~$y \in L$)} \\
  \sigma_{xy} \wedge \sigma_{xy'} &\vdash \bigvee\{ \top \,|\, y = y' \} & \text{(for each~$x,y,y' \in L$)} \\
  \sigma_{xy} \wedge \sigma_{x'y} &\vdash \bigvee\{ \top \,|\, x = x' \} & \text{(for each~$x,x',y \in L$)} \\
  \top &\vdash \sigma_{00} \\
  \top &\vdash \sigma_{11} \\
  \sigma_{xy} \wedge \sigma_{x'y'} &\vdash \sigma_{x+x',y+y'} & \text{(for each~$x,x',y,y' \in L$)} \\
  \sigma_{xy} \wedge \sigma_{x'y'} &\vdash \sigma_{xx',yy'} & \text{(for each~$x,x',y,y' \in L$)} \\
  \top &\vdash \sigma_{xx} & \text{(for each~$x \in k$)}
\end{align*}
A model of this theory (can be reorganized to be) an element of the classical
Galois group of~$L|k$.

\paragraph{The inconsistent geometric theory.} This theory has no sorts,
function symbols or relations, but the single axiom $\top \vdash \bot$.
This theory does not admit any set-based models.

\paragraph{The empty geometric theory.} This theory does not have any sorts,
function symbols, relations or axioms. There is exactly one set-based model of
this theory, the empty structure.

@@

\begin{exercise}{Discrete locales as classifying locales}%
\begin{alphlist}[(b)]
\item Find a theory which is classified by the two-point locale (the locale
associated to the discrete two-point topological space).
\item Let~$M$ be a set and endow it with the discrete topology. Find a theory
which is classified by~$L(M)$.
\end{alphlist}
\end{exercise}

\begin{exercise}{The localic unit interval}%
Extend the theory of Dedekind cuts in such a way that its models are those
Dedekind reals which are elements of~$[0,1]$.
\end{exercise}

>>>>

% NOTE TO EDITORS AND LAYOUTERS: This package increases the quality of the
% typesetting. I personally recommebd using it, but my contribution does not
% depend on it.
%\usepackage[expansion=true,protrusion=true]{microtype}

@@

    \textnormal{mixed rules (unnecessary in presence of the double rule for implication)} \\\smallskip
    \vspace{-0.5em}
    \phantom{a}\hfill
    \AxiomC{$\phantom{\seq{\vec x}}$}
    \UnaryInfC{$(\bigvee_i \varphi_i) \wedge \psi \seq{\vec x} \bigvee_i (\varphi_i \wedge \psi)$}
    \DisplayProof
    \hfill
    \AxiomC{$\phantom{\seq{\vec x}}$}
    \UnaryInfC{$(\exists y\?Y\_ \varphi) \wedge \psi \seq{\vec x} \exists y\?Y\_ (\varphi \wedge \psi)$}
    \DisplayProof
    \makebox[0pt]{\shifttext{\tiny ($y$ none of the~$\vec x$)}}
    \hfill\phantom{a}
    \bigskip

@@

We explicitly allow empty carriers, that is we do not demand that the
sets~$\brak{X}$ are inhabited.
Examples for geometric theories include the following.
\begin{enumerate}
\item \emph{The geometric theory of rings.} This theory has one sort, $R$; five function symbols:
$0$ and $1$ (nullary), $-$ (unary), $+$ and $\cdot$ (binary); no relation
symbols; and the usual axioms, such as $\top \vdash_{x:R,y:R} x + y = y + x$.

@@

In a similar vein, there is the geometric theory of monoids, of groups, and so
on. Some of these theories use the infinite disjunctions supported by geometric
logic: For instance, the theory of fields of positive characteristic or the
theory of torsion groups have among their axioms
\[ \top \vdash \bigvee_{n \in \NN_{\geq 1}} (\underbrace{1 + \cdots + 1}_{\text{$n$ summands}} = 0)
  \qquad
  \top \seq{g:G} \bigvee_{n \in \NN} (g^n = e). \]

\item \emph{The geometric theory of objects.}\label{item:theory-of-objects}
This theory has one sort, $X$, and no function
symbols, relation symbols or axioms. A set-based model of this theory is just a
set.

@@

\item \emph{The geometric theory of surjections~$f : \NN \twoheadrightarrow
\RR$.} This theory has no
sorts; no function symbols; a nullary relation symbol~$\varphi_{nx}$ for each
pair~$\langle n,x \rangle \in \NN \times \RR$ (to be read as~``$f$ maps~$n$
to~$x$''); and the following axioms:
\begin{enumerate}
\item For each~$n \in \NN$, the axiom~$\top \vdash \bigvee_{x \in \RR}
\varphi_{nx}$, expressing that~$f$ is total.
\item For each~$n \in \NN$ and each~$x,y \in \RR$, the axiom~$\varphi_{nx}
\wedge \varphi_{ny} \vdash \bigvee\{ \top \,|\, x = y \}$,
expressing that~$f$ is single-valued.\footnote{The disjunction is taken over
the set~$\{ \top \,|\, x = y \}$. This is a certain subsingleton set of
formulas; it is inhabited (by the formula~$\top$) iff~$x = y$. Hence, if~$x =
y$, this axiom reads~$\varphi_{nx}
\wedge \varphi_{ny} \vdash \top$ and could also be omitted; if~$x \neq y$,
this axiom reads~$\varphi_{nx} \wedge \varphi_{ny} \vdash \bot$.}
\item For each~$x \in \RR$, the axiom~$\top \vdash \bigvee_{n \in \NN}
\varphi_{nx}$, expressing that~$f$ is surjective.
\end{enumerate}

@@

\item \emph{The geometric theory of Dedekind cuts.}\label{par:dedekind-cuts}
This theory has no sorts; no
function symbols; nullary relation symbols~$\alpha_x$ and~$\beta_x$ for each
rational number~$x$ (to be read as~``$x$ is contained in the lower (upper)
cut''); and the following axioms, spelling out that the cut is bounded,
rounded, open and located:
\begin{align*}
  \top &\vdash \textstyle\bigvee_{x \in \QQ} \alpha_x \\
  \top &\vdash \textstyle\bigvee_{x \in \QQ} \beta_x \\
  \alpha_y &\vdash \alpha_x & \text{(for each~$x,y \in \QQ$ such that~$x < y$)} \\
  \beta_x &\vdash \beta_y & \text{(for each~$x,y \in \QQ$ such that~$x < y$)} \\
  \alpha_x &\vdash \textstyle\bigvee_{y > x} \alpha_y & \text{(for each~$x \in \QQ$)} \\
  \beta_y &\vdash \textstyle\bigvee_{x < y} \beta_x & \text{(for each~$y \in \QQ$)} \\
  \alpha_x \wedge \beta_y &\vdash \bot & \text{(for each~$x,y \in \QQ$ such that~$x \geq y$)} \\
  \top &\vdash \alpha_x \vee \beta_y & \text{(for each~$x,y \in \QQ$ such that~$x < y$)}
\end{align*}

@@

\item \emph{The geometric theory of prime ideals of a given ring~$A$.} In classical
commutative algebra, the notion of a \emph{prime ideal} of a ring~$A$ is
fundamental. Corresponding to the definition (recalled in
Section~\ref{sect:algebraic-preliminaries}), the geometric theory of prime
ideals has no sorts; no function
symbols; one relation symbol~``$V(x)$'' for each element~$x \in A$; and the
following axioms:
\begin{align*}
  \top &\vdash V(0) \\
  V(x) \wedge V(y) &\vdash V(x+y) & \text{(for each~$x,y \in A$)} \\
  V(x) &\vdash V(xy) & \text{(for each~$x,y \in A$)} \\
  V(1) &\vdash \bot \\
  V(xy) &\vdash V(x) \vee V(y) & \text{(for each~$x,y \in A$)}
\end{align*}

@@

\item \emph{The geometric theory of automorphisms of a field.}\label{par:theory-automorphisms}
Let~$L|k$ be a field
extension. The classifying locale of automorphisms~$L \to L$ which fix~$k$ has
one nullary relation symbol~``$\sigma_{xy}$'' for each pair~$\langle x,y
\rangle \in L$ and the following axioms:
\begin{align*}
  \top &\vdash \bigvee_y \sigma_{xy} & \text{(for each~$x \in L$)} \\
  \top &\vdash \bigvee_x \sigma_{xy} & \text{(for each~$y \in L$)} \\
  \sigma_{xy} \wedge \sigma_{xy'} &\vdash \bigvee\{ \top \,|\, y = y' \} & \text{(for each~$x,y,y' \in L$)} \\
  \sigma_{xy} \wedge \sigma_{x'y} &\vdash \bigvee\{ \top \,|\, x = x' \} & \text{(for each~$x,x',y \in L$)} \\
  \top &\vdash \sigma_{00} \\
  \top &\vdash \sigma_{11} \\
  \sigma_{xy} \wedge \sigma_{x'y'} &\vdash \sigma_{x+x',y+y'} & \text{(for each~$x,x',y,y' \in L$)} \\
  \sigma_{xy} \wedge \sigma_{x'y'} &\vdash \sigma_{xx',yy'} & \text{(for each~$x,x',y,y' \in L$)} \\
  \top &\vdash \sigma_{xx} & \text{(for each~$x \in k$)}
\end{align*}
A model of this theory (can be reorganized to be) an element of the classical
Galois group of~$L|k$.

\item \emph{The inconsistent geometric theory.} This theory has no sorts,
function symbols or relations, but the single axiom $\top \vdash \bot$.
This theory does not admit any set-based models.

\item \emph{The empty geometric theory.} This theory does not have any sorts,
function symbols, relations or axioms. There is exactly one set-based model of
this theory, the empty structure.
\end{enumerate}

\paragraph{Extracting constructive content from classical proofs.} Geometric
logic strikes a fine balance between being sufficiently expressive in order to
have rich applications, but also sufficiently restrictive so that \emph{Barr's
theorem} holds. This theorem comes in two versions, both concerned with
eliminating classical reasoning from proofs of geometric sequents:
\begin{enumerate}
\item If infinitary classical logic proves a geometric sequent, then so does
geometric logic.
\item If a specific version of extensional type theory with the law of excluded
middle and the axiom of choice proves a geometric sequent, then so does
geometric logic.
\end{enumerate}
For instance, consider the theory of algebraically closed geometric fields,
that is the theory of rings extended by the axioms
\begin{align*}
  \top &\seq{x:R} (x = 0) \vee (\exists y\?R\_ xy = 1) \\
  \top &\seq{a_0:R,\ldots,a_{n-1}:R} \exists x\?R\_ x^n + a_{n-1}x^{n-1} +
  \cdots + a_1x + a_0 = 0.
\end{align*}
The \emph{Nullstellensatz} is the statement that for any
system~$f_1,\ldots,f_n \in K[X_1,\ldots,X_m]$ of polynomials over an
algebraically closed field~$K$, either the polynomials~$f_i$ have a common zero
or there are polynomials~$g_i$ such that~$1 = \sum_i f_i g_i$ (such an
expression is an \emph{algebraic witness} of the nonexistence of a common
zero).

The Nullstellensatz is a theorem of classical mathematics, employing the
axiom of choice. Since it can be formulated as a geometric sequent over the
geometric theory of algebraically closed geometric fields, Barr's theorem
implies that there is also a constructive proof, in fact even a purely
geometric proof.

The simpler version of Barr's theorem is constructive, and although we will not
use it in the remainder of this text, we take the opportunity to give a
self-contained proof.

However, the more powerful version is not constructive. This puts us in the
curious situation that we are promised, unconstructively, that there exists
a constructive proof. The textbook argument~\cite[Theorem~7.57]{johnstone:topos-theory} establishing the more powerful
version of Barr's theorem requires
the axiom of choice, but recently Michael Rathjen showed that~\textsc{zf}
suffices~\cite{rathjen:barr}.

\begin{theorem}Let~$\TT$ be a geometric theory. Let~$\sigma$ be a geometric
sequent over the signature of~$\TT$. Then the following statements are
equivalent.
\begin{enumerate}
\item There is a~$\TT$-derivation of~$\sigma$ in geometric logic.
\item There is a~$\TT$-derivation of~$\sigma$ in infinitary intuitionistic logic.
\item There is a~$\TT$-derivation of~$\sigma$ in infinitary classical logic.
\end{enumerate}
\end{theorem}

\begin{proof}It is immediate that~(1) implies~(3).

To show that~(3) implies~(2), we combine the double negation translation
with Friedman's trick. We give a sketch of the main thrust of the argument,
omitting lengthy verifications. Let~$\sigma \equiv (\alpha \seq{\vec x} \beta)$.
We introduce the operator~$\nabla$ with~$\nabla\varphi \defeqv
((\varphi \Rightarrow \beta) \Rightarrow \beta)$, where care has to be taken to
avoid capturing the variables~$\vec x$. This operator is a \emph{local
operator} in the sense that for any formula~$\varphi$ in some context~$\vec y$,
infinitary intuitionistic logic proves the sequents
\[
\top \seq{\vec x,\vec y} (\varphi \Rightarrow \nabla\varphi),\
\top \seq{\vec x,\vec y} (\nabla\nabla\varphi \Rightarrow \nabla\varphi),\
\top \seq{\vec x,\vec y} (\nabla\varphi \wedge \nabla\psi \Leftrightarrow \nabla(\varphi \wedge \psi)).
\]
We then set up the~\emph{$\nabla$-translation}~$\varphi \mapsto \varphi^\nabla$
for infinitary first-order formulas over the signature of~$\TT$ as described in
Table~\ref{table:nabla-translation}.
Again, care has to be taken to ensure that there is no variable collision with the
variables~$\vec x$. The chosen local operator is special in that
the~$\nabla$-translation of the law of excluded middle,
\[ \top \seq{\vec x,\vec y} \nabla(\varphi^\nabla \vee (\varphi^\nabla
\Rightarrow \nabla\bot)) \]
is intuitionistically provable. Straightforward inductions on the structure of
formulas or derivations then establish the following properties.
\begin{alphlist}
\item For any infinitary first-order formula~$\varphi$ in some context~$\vec y$,
infinitary intuitionistic logic proves~$\nabla(\varphi^\nabla) \seq{\vec x,\vec
y} \varphi^\nabla$.
\item For any geometric formula~$\varphi$ in some context~$\vec y$, infinitary
intuitionistic logic proves~$\varphi^\nabla \mathrel{\dashv\vdash\!\!\!_{\vec x,\vec y}} \nabla\varphi$.
\item For any derivation of a sequent~$\varphi \seq{\vec y} \psi$ in infinitary
classical logic, there is a derivation of~$\varphi^\nabla \seq{\vec y}
\psi^\nabla$ in infinitary intuitionistic logic.
\end{alphlist}
Combining these ingredients, the given derivation of~$\sigma \equiv (\alpha
\seq{\vec x} \beta)$ in infinitary classical logic can be transformed to give a derivation of
\begin{align*}
  \alpha[\vec x'/\vec x] &\seq{\vec x,\vec x'}
  \nabla(\alpha[\vec x'/\vec x]) \seq{\vec x,\vec x'}
  (\alpha[\vec x'/\vec x])^\nabla \\
  &\seq{\vec x,\vec x'}
  (\beta[\vec x'/\vec x])^\nabla \seq{\vec x,\vec x'}
  \nabla(\beta[\vec x'/\vec x]) \mathrel{\dashv\vdash\!\!\!_{\vec x,\vec x'}}
  ((\beta[\vec x'/\vec x] \Rightarrow \beta) \Rightarrow \beta).
\end{align*}
in infinitary intuitionistic logic. Substituting~$\vec x'$ by~$\vec x$, we obtain
\[ \alpha \seq{\vec x} ((\beta \Rightarrow \beta) \Rightarrow \beta) \seq{\vec
x} \beta. \]

To verify that~(2) implies~(1), XXX
\end{proof}

\begin{table}[ht]
  \tbl{The definition of the~$\nabla$-translation}{\begin{minipage}{\textwidth}\begin{align*}
    (x = y)^\nabla &\defeqv \nabla(x = y) \\
    R(y_1,\ldots,y_m)^\nabla &\defeqv \nabla(R(y_1,\ldots,y_m)) \\
    \top^\nabla &\defeqv \nabla\top \quad \\
    \bot^\nabla &\defeqv \nabla\bot \\
    (\varphi \wedge \psi)^\nabla &\defeqv \varphi^\nabla \wedge \psi^\nabla &
    \textstyle (\bigwedge_i \varphi_i)^\nabla &\defeqv \textstyle \bigwedge_i \varphi_i^\nabla \\
    (\varphi \vee \psi)^\nabla &\defeqv \nabla(\varphi^\nabla \vee \psi^\nabla) &
    \textstyle (\bigvee_i \varphi_i)^\nabla &\defeqv \textstyle \nabla(\bigvee_i \varphi_i^\nabla) \\
    (\varphi \Rightarrow \psi)^\nabla &\defeqv (\varphi^\nabla \Rightarrow \psi^\nabla) \\
    (\forall x\?X\_ \varphi)^\nabla &\defeqv (\forall x\?X\_ \varphi^\nabla) &
    (\exists x\?X\_ \varphi)^\nabla &\defeqv \nabla(\exists x\?X\_ \varphi^\nabla)
  \end{align*}\end{minipage}}
  \label{table:nabla-translation}
\end{table}

@@

\begin{exercise}{The theory of finite objects}%
\begin{alphlist}[(c)]
\item Extend the geometric theory of objects
(page~\ref{item:theory-of-objects}) by relation symbols and axioms, but not by
constant symbols, to ensure that the set-based models are precisely the
two-element sets.
\item Write down a geometric theory whose set-based models are precisely the
(Kuratowski-)finite sets.\smallskip

{\scriptsize\emph{Hint.} Employ countably many relation symbols~$R_n
\hookrightarrow X \times \cdots \times X$, where~$R_n(x_1,\ldots,x_n)$ should
signify that the elements~$x_1,\ldots,x_n$ exhaust all elements of~$X$. A
solution is given in~\cite[Example~D1.1.7(k)]{johnstone:elephant}.\par}
\end{alphlist}
\end{exercise}

\begin{exercise}{Discrete locales}%
\begin{alphlist}[(b)]
\item Find a theory which is classified by the two-point locale (the locale
associated to the discrete two-point topological space).
\item Let~$M$ be a set and endow it with the discrete topology. Find a theory
which is classified by~$L(M)$.
\end{alphlist}
\end{exercise}

\begin{exercise}{The localic unit interval}%
Extend the theory of Dede\-kind cuts in such a way that its models are those
Dedekind reals which are elements of~$[0,1]$.
\end{exercise}
